Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-26 Manuscript/manuscript.Rmd
added supp legends at bottom

Genomic annotation data in the form of cell type specific chromatin states, histone marks, and transcription factor binding sites (GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Tables 2-3). Functional annotations were represented by chromosome-specific regions on the linear genome obtained by ChIP-seq peak calling algorithms.

@@

A total of 78 cell type specific functional genomic annotations for the GM12878 and K562 cell line were used to build the feature space of downstream predictive models (Supplementary Tables 2-3). The annotations could be broken down into three sets including histone modifications (HM), chromatin state segmentations (ChromHMM), and transcription factor binding sites (TFBS). We constructed three different predictor types for each functional genomic annotation (Figure 1B). Counts of overlaps (**overlap count**, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (**overlap percent**, OP) were defined as the proportion of a bin overlapped by an annotation region. **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of distances (Supplementary Figures 4-5). These feature engineering steps were performed to improve TAD boundary predictions by accounting for direct (OC), proportional (OP), and indirect (distance) relationships between TAD boundaries and genomic annotations.

@@

We found that, regardless of resampling technique, random forest classifiers built using a distance-type feature space yielded consistently higher MCCs when compared to the overlap count and overlap percent predictor types (Figure 3), and these results were consistent across resolutions and cell types (Supplementary Figure 6). This indicated that models are better able to discriminate between TAD and non-TAD boundary regions when provided with spatially related associations.

@@

We then aimed to compare the performances of annotation-specifc models. In doing so, we inspected the proportion of predictive importance that was explained by each of the three groups of genomic annotations (histone modifications (HM), chromatin states (ChromHMM), and transcription factor binding sites (TFBS)). For optimal random forest classifiers, we found that the majority of the proportional predictive importance was obtained by transcription factor binding sites (Supplementary Figure 7). Furthermore, models build solely on TFBSs outperformed other annotation-specific models at all resolutions (Figure 4A; Supplementary Figure 8A). Notably, the TFBS-specific models outperformed models built on the combined sets of annotations, and this effect was most pronounced at lower resolutions. This indicated that the inclusion of other types of annotations contributed to additional noise rather than signal in TAD boundary prediction. In summary, these results suggest that TFBS are the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

@@

Given that many TFBS are known to colocalize together on the linear genome and are therefore highly correlated (Supplementary Figure 9), we sought to avoid overfitting by performing recursive feature elimination. We found that our models were able to obtain near-optimal performance using approximately eight TFBSs (Figure 4B; Supplementary Figure 8B). However, given that we train our models on chromosome-specific data, the most significant annotations vary for each chromosome. To determine the most common TFBSs, we clustered the predictive importance values across chromosomes (Figure 4C; Supplementary Figure 8C). We found that, similar to other studies [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD boundary regions.

@@

After obtaining our optimally built random forest classifier, we investigated whether our predictive models can alleviate resolution limitations of Hi-C data by predicting TAD boundaries at base pair resolution. We were able to identify regions where the model predicted the base pair to be a TAD boundary with probability equal to 1. [??? Where is this figure? It seemed important] By clustering the regions (see Methods), we were able to predict a comparable number of TAD boundary regions as called by the ARROWHEAD TAD caller. However, only 24% and 29% of predicted and called (and 10 kb flanked) TAD boundaries overlapped for the GM12878 and K562 cell line, respectively, suggesting differing locations in TAD boundaries throughout the linear genome (Supplementary Figure 10). [??? Need further drill down, as we discussed]

For each predicted TAD boundary region, we were able to identify the summit locations of TAD boundary location by performing partitioning around medoids (see Methods). To compare the summit locations of predicted and called TAD boundaries, we evaluated the signal strength of ChIP-seq peak regions around a 10 kb flanking region. We found that the signal strength for the CTCF, RAD21, SMC3, and ZNF143 [??? Make all in this order!] TFBS peaked at our predicted TAD boudaries, while was poorly defined around the called TAD boundaries (Figure 5; Supplementary Figure [???]).  Thus, our model is able to alleviate resolution limitations and precisely locate TAD boundaries that correspond to the highest signal strength of known TAD boundary-forming TFBSs.

@@

>>>>

Genomic annotation data in the form of cell type specific chromatin states, histone marks, and transcription factor binding sites (GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 2). Functional annotations were represented by chromosome-specific regions on the linear genome obtained by ChIP-seq peak calling algorithms.

@@

A total of 78 cell type specific functional genomic annotations for the GM12878 and K562 cell line were used to build the feature space of downstream predictive models (Supplementary Table 2). The annotations could be broken down into three sets including histone modifications (HM), chromatin state segmentations (ChromHMM), and transcription factor binding sites (TFBS). We constructed three different predictor types for each functional genomic annotation (Figure 1B). Counts of overlaps (**overlap count**, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (**overlap percent**, OP) were defined as the proportion of a bin overlapped by an annotation region. **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of distances (Supplementary Figure 4). These feature engineering steps were performed to improve TAD boundary predictions by accounting for direct (OC), proportional (OP), and indirect (distance) relationships between TAD boundaries and genomic annotations.

@@

We found that, regardless of resampling technique, random forest classifiers built using a distance-type feature space yielded consistently higher MCCs when compared to the overlap count and overlap percent predictor types (Figure 3), and these results were consistent across resolutions and cell types (Supplementary Figure 5). This indicated that models are better able to discriminate between TAD and non-TAD boundary regions when provided with spatially related associations.

@@

We then aimed to compare the performances of annotation-specifc models. In doing so, we inspected the proportion of predictive importance that was explained by each of the three groups of genomic annotations (histone modifications (HM), chromatin states (ChromHMM), and transcription factor binding sites (TFBS)). For optimal random forest classifiers, we found that the majority of the proportional predictive importance was obtained by transcription factor binding sites (Supplementary Figure 6). Furthermore, models build solely on TFBSs outperformed other annotation-specific models at all resolutions (Figure 4A; Supplementary Figure 7A). Notably, the TFBS-specific models outperformed models built on the combined sets of annotations, and this effect was most pronounced at lower resolutions. This indicated that the inclusion of other types of annotations contributed to additional noise rather than signal in TAD boundary prediction. In summary, these results suggest that TFBS are the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

@@

Given that many TFBS are known to colocalize together on the linear genome and are therefore highly correlated (Supplementary Figure 8), we sought to avoid overfitting by performing recursive feature elimination. We found that our models were able to obtain near-optimal performance using approximately eight TFBSs (Figure 4B; Supplementary Figure 7B). However, given that we train our models on chromosome-specific data, the most significant annotations vary for each chromosome. To determine the most common TFBSs, we clustered the predictive importance values across chromosomes (Figure 4C; Supplementary Figure 7C). We found that, similar to other studies [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD boundary regions.

@@

After obtaining our optimally built random forest classifier, we investigated whether our predictive models can alleviate resolution limitations of Hi-C data by predicting TAD boundaries at base pair resolution. We were able to identify regions where the model predicted the base pair to be a TAD boundary with probability equal to 1. [??? Where is this figure? It seemed important] By clustering the regions (see Methods), we were able to predict a comparable number of TAD boundary regions as called by the ARROWHEAD TAD caller. However, only 24% and 29% of predicted and called (and 10 kb flanked) TAD boundaries overlapped for the GM12878 and K562 cell line, respectively, suggesting differing locations in TAD boundaries throughout the linear genome (Supplementary Figure 9). [??? Need further drill down, as we discussed]

For each predicted TAD boundary region, we were able to identify the summit locations of TAD boundary location by performing partitioning around medoids (see Methods). To compare the summit locations of predicted and called TAD boundaries, we evaluated the signal strength of ChIP-seq peak regions around a 10 kb flanking region. We found that the signal strength for the CTCF, RAD21, SMC3, and ZNF143 [??? Make all in this order!] TFBS peaked at our predicted TAD boudaries, while was poorly defined around the called TAD boundaries (Figure 5; Supplementary Figure 10[???; !!!fixed]).  Thus, our model is able to alleviate resolution limitations and precisely locate TAD boundaries that correspond to the highest signal strength of known TAD boundary-forming TFBSs.

@@

# Supplementary Legends

**Supplementary Table 1. Hi-C data.** A complete list of the Hi-C data libraries used to obtain the locations of topologically associating domains.

**Supplementary Table 2. Genomic annotations.** A complete list of the cell line specific genomic annotations used to build the feature space for all downstream models. The GRCh37/hg19 human genome assembly was used. "Genomic Class" - broad category of genomic features, "Element" - names of genomic annotations, "Cell Type-Specific Source" - download URL specific to the cell line (not all annotations were provided by the same institutions).

**Supplementary Figure 1. Diagram of the number of called TADs and TAD boundaries.** TADs were called using the ARROWHEAD pipeline [@durand2016juicebox] at 5 resolutions: 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb, for both the GM12878 and K562 cell line, across each autosomal chromosome (excluding chromosome 9). A preprocessing step was included to filter the TADs that exceeded 2 mb in width. The coordinates of unique TAD boundaries were then extracted and enumerated at each resolution.

**Supplementary Figure 2. Venn diagrams showing that TAD boundary regions exhibit cell-line specificity across resolution.**

**Supplementary Figure 3. Boxplots illustrating that the cell line specific proportion of class imbalance remained severe irrespective of resolution.** Proportion of imbalance (PI) was defined as the ratio of TAD boundary regions to the sum of both TAD and non-TAD boundary regions. Balanced datasets have PI close to 0.50.

**Supplementary Figure 4. Performing $log_{2}$ transformations normalizes skewed distance densities across resolution.** When measured in basepairs, distances were highly right skewed (red lines) across all resolutions for both the GM12878 (A) and K562 (B) cell lines. After $log_{2}$ transforming the distances (blue lines), densities became much more gaussian distributed for all resolutions.

**Supplementary Figure 5. Models built using distance-type predictors combined with SMOTE re-sampling yielded optimal MCCs across resolution on K562.** Barplots comparing the performances of random forest classifiers built using different combinations of data level characteristics for K562. Within each re-sampling technique (A-D), performances are compared for each type of feature space: OC (red), OP (green), and Distance (blue), across resolutions. Performances were evaluated using MCCs and aggregated by averaging across each chromosome, with error bars representing 1 standard deviation away from the mean.

**Supplementary Figure 6. The proportional importance of optimally built cell line specific models was outweighed by transcription factor binding sites across resolutions.** Stacked barplots illustrate the proportion of total predictive importance in cell line specific random forest classifiers for (A) GM12878 and (B) K562 respectively. Model inputs included the optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors). Predictive importance was averaged for models built on chromosome specific regions of the linear genome. Proportional importance was then calculated according to the annotation group of the feature space including histone modifications (HM), chromatin state segmentations (ChromHMM), and transcription factor binding sites (TFBS).

**Supplementary Figure 7. Assessing top ranked transcription factors in TAD boundary prediction for K562.** (A) Barplots comparing performances of models using K562-specific groups of functional genomic annotation as the feature space including histone modifications (HM; red), chromatin states (ChromHMM; green), transcription factor binding sites (TFBS; blue), in addition to a model with a feature space containing all 3 groups (Combined; purple). Error bars represent ± 1 standard deviation from the mean MCC across each chromosomal section of the linear genome. (B) Recursive feature elimimation analysis used to reduce the number of TFBS to include in the optimally build random forest classifier. Error bars represent 1 standard deviation away from the mean cross-validated MCC. (C) Clustered heatmap of variable importances for the union of the top 8 most predictive TFBS across chromosomes.

**Supplementary Figure 8. Pairwise Jaccard Indices illustrate correlations between cell line specific transcription factor binding sites for GM12878 (A) and K562 (B).**

**Supplementary Figure 9. Venn diagrams illustrating the amount of overlap between predicted and called TAD boundary regions.** For each of the GM12878 (A) and K562 (B), both predicted and called TAD boundaries were flanked other either side by 10 kb and the amount of overlap was enumerated.

**Supplementary Figure 10. Predicted TAD boundaries were centered closer to peak signal regions defined by known architectural transcription factors when compared to called TAD boundaries for K562.** Enrichment
heatmaps illustrate the average signal strength of K562-specific ChIP-seq peak regions defined by (A) SMC3, (B) ZNF143, (C) CTCF, and (D) RAD21 around the centers of 10 kb flanked predicted and called TAD boundaries.
